---
title: "Library kit as a nuisance factor"
format:
  html:
    theme: yeti
    code_folding: hide
    self_contained: yes
    code-fold: true
    code-tools: true
editor: visual
date: today
author: "Joowook Kim"
execute:
  warning: false
params:
  project_name: "Library kit as a nuisance factor"
---

# Load lib

```{r}
suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(ggplot2)
  library(edgeR)
  library(DESeq2)
  library(limma)
  library(pheatmap)
  library(RColorBrewer)
  library(ComplexHeatmap)
  library(ExploreModelMatrix)
})

```

# Experiment design

2 factor design with 3 replicates:

-   treatment factor ( IL-1$\beta$ treated/untreated), 2 levels

-   nuisance factor (libraryKit - v4kit(Takara)/Pico(Takara)/truseq(illumina)), 3 levels

```{r, Load data}
v4 <- read_tsv("https://ftp.ncbi.nlm.nih.gov/geo/series/GSE124nnn/GSE124167/suppl/GSE124167_FINAL_master_list_of_gene_counts_MIN.V4.txt.gz") |>
  dplyr::select(-geneCoordinate, -geneSymbol)

pico <- read_tsv("https://ftp.ncbi.nlm.nih.gov/geo/series/GSE124nnn/GSE124167/suppl/GSE124167_FINAL_master_list_of_gene_counts_MIN.sense.Pico.R.txt.gz") |>
  dplyr::select(-geneCoordinate, -geneSymbol)

illum <- read_tsv("https://ftp.ncbi.nlm.nih.gov/geo/series/GSE124nnn/GSE124167/suppl/GSE124167_FINAL_master_list_of_gene_counts_MIN.sense.Truseq.txt.gz") |>
  dplyr::select(-geneCoordinate, -geneSymbol)

annot_df <- read_tsv("https://ftp.ncbi.nlm.nih.gov/geo/series/GSE124nnn/GSE124167/suppl/GSE124167_FINAL_master_list_of_gene_counts_MIN.sense.Truseq.txt.gz") |>
  dplyr::select(id, geneSymbol)

colnames(illum) <- sub("^Illumina\\.(ILB|UNT)_(\\d+)$", "\\1_\\2_truseq", colnames(illum))

## merge data frames
merged_df <- Reduce(function(x, y) merge(x, y, by = "id", all = TRUE), list(v4, pico, illum))

## clean column name
colnames(merged_df) <- colnames(merged_df) |> janitor::make_clean_names()

## remove clonthe from colnames
colnames(merged_df) <- colnames(merged_df) |> stringr::str_remove("_clontech")


## check NAs
print("checking NAs")
is.na(merged_df) |> sum()

## replace NAs with 0
merged_df[is.na(merged_df)] <- 0

count_df <- merged_df
rm(merged_df)
```

## Meta data

```{r, Meta data}

## remove 'id' in colnames
col <- colnames(count_df)[-1]
meta <- data.frame(
  sample = col,
  ilb_trt = ifelse(grepl("^ilb_", col), "treated", "untreated"),
  kit = sub(".*_(truseq|pico|v4)$", "\\1", col),
  stringsAsFactors = FALSE
) |>
  mutate(ilb_trt = as.factor(ilb_trt),
         kit = as.factor(kit)
         )

rm(col)

DT::datatable(meta, filter = 'top')

## change reference level for factors
meta$ilb_trt <- relevel(meta$ilb_trt, "untreated")
meta$kit <- relevel(meta$kit, "truseq")

## check reference levels for factors
print("chekcing reference levels ")
str(meta)
```

# EDA

## Before the kit effect adjusted

The PCA plot shows substantial differences between library kits — PC1 explains 90% of the variance and appears to separate the different kit samples, while PC2 seems to separate between the treated and untreated.

```{r, EDA1}

rownames(count_df) <- count_df$id
count_df <- count_df |> dplyr::select(-id)

rownames(meta) <- meta$sample
meta <- meta |> dplyr::select(-sample)

## sanity check!!
all(rownames(meta) %in% colnames(count_df))

## this is for EDA
dds <- DESeqDataSetFromMatrix(countData = count_df,
                              colData = meta,
                              design = ~ 1)

## filter gene with less than 10 reads
dds <- dds[ rowSums(counts(dds)) > 10, ]

rld <- rlog(dds, blind=FALSE)

pca_helper <- function(object, ntop=500, meta, sample_name = "sample") {
  # object is either vst or rlog

  # calculate the variance for each gene
  # rv <- rowVars(assay(object))
  rv <- rowVars(object)

  # select the ntop genes by variance
  select <- order(rv, decreasing=TRUE)[seq_len(min(ntop, length(rv)))]

  # perform a PCA on the data in assay(x) for the selected genes
  pca_obj <- prcomp(t((object)[select,]))

  # get pca scores
  pca_scores <- pca_obj$x %>% data.frame()

  # get eigen values
  pca_eigen_val <- round((summary(pca_obj)$importance[2,] * 100),1)
  # return (pca)

  # merge pca_socres and meta_df
  merged_df <- pca_scores |> tibble::rownames_to_column(sample_name)
  pca_merged_df <- left_join(merged_df, meta |>
                               tibble::rownames_to_column(sample_name), by = sample_name)

  result <- list(pca_merged_df, pca_eigen_val)
  names(result) <- c("pca_merged_df", "pca_eigen_val")

  return(result)
}

pca_merged_list <- pca_helper(assay(rld), ntop = Inf, meta = meta , sample_name = 'sample')
## plot pca
## pca pc1/pc2
# see https://stackoverflow.com/questions/56547256/show-3-factors-ggplot-geom
pcaplot_pc1_pc2 <- ggplot(pca_merged_list$pca_merged_df, aes(x=PC1, y=PC2, color=kit, shape=ilb_trt)) +
  geom_point(size=5.5) +
  # coord_fixed() +
  # scale_size_manual(values=c(7,5))+
  ggrepel::geom_text_repel(label=pca_merged_list$pca_merged_df$sample, box.padding = .5) +
  xlab(paste0("PC1: ", pca_merged_list$pca_eigen_val[1], "%")) +
  ylab(paste0("PC2: ", pca_merged_list$pca_eigen_val[2], "%")) +
  ggtitle("PCA plot using all genes w rlog")  +
  theme_bw() +
  theme(aspect.ratio=1) +
  theme(text = element_text(size = 30)) +
  scale_size_manual(values = c(8,5))

print(pcaplot_pc1_pc2)

## pca pc1/pc3
# pcaplot_pc1_pc3 <- ggplot(pca_merged_df, aes(x=PC1, y=PC3, color=cancer_subtype, shape=batch)) +  #, shape=Treat, size=Sex)) +
#   geom_point(size=5.5) +
#   # coord_fixed() +
#   ggrepel::geom_text_repel(label=pca_merged_df$sample_name, box.padding = .5) +
#   xlab(paste0("PC1: ", pca_eigen_val[1], "%")) +
#   ylab(paste0("PC3: ", pca_eigen_val[3], "%")) +
#   ggtitle("PCA plot using all genes")  +
#   theme_bw() +
#   theme(aspect.ratio=1) +
#   theme(text = element_text(size = 30))

# print(pcaplot_pc1_pc3)

sampleDists <- dist( t( assay(rld) ) )

sampleDistMatrix <- as.matrix( sampleDists )

annotation_col = data.frame(
    ilb_trt = meta$ilb_trt,
    kit = meta$kit
)

heat_anno <- HeatmapAnnotation(
  df=annotation_col
)

set.seed(0)
hmap <- pheatmap(sampleDistMatrix,
                # clustering_distance_columns = sampleDists,
                # clustering_distance_rows = sampleDists,
                # rect_gp = gpar(col = "white", lwd = 2),
                # # row_names_gp = gpar(fontsize = 5),
                # # row_dend_side = 'left',
                # column_title = "Heatmap of the sample-to-sample based on Euclidean distances",
                heatmap_legend_param = list(
                                title = "Euclidean dist"),
                annotation_col = annotation_col
                )

print(hmap)
```

## After the kit effect adjusted

```{r, EDA2}
rld_filt <- rld
assay(rld_filt)<- limma::removeBatchEffect(assay(rld_filt), rld_filt$kit)

pca_merged_list <- pca_helper(assay(rld_filt), ntop = Inf, meta = meta , sample_name = 'sample')

pcaplot_pc1_pc2 <- ggplot(pca_merged_list$pca_merged_df, aes(x=PC1, y=PC2, color=kit, shape=ilb_trt)) +
  geom_point(size=5.5) +
  # coord_fixed() +
  # scale_size_manual(values=c(7,5))+
  ggrepel::geom_text_repel(label=pca_merged_list$pca_merged_df$sample, box.padding = .5) +
  xlab(paste0("PC1: ", pca_merged_list$pca_eigen_val[1], "%")) +
  ylab(paste0("PC2: ", pca_merged_list$pca_eigen_val[2], "%")) +
  ggtitle("PCA plot using all genes w rlog")  +
  theme_bw() +
  theme(aspect.ratio=1) +
  theme(text = element_text(size = 15)) +
  scale_size_manual(values = c(8,5))

print(pcaplot_pc1_pc2)


```

# Fit a model to the data

There are different ways of writing out/parameterization of model. One is `cell means model` and another is `effect model` and the other is `reference coded/mean reference model`. The reference coded model is R's default setting.

> The reference coded model writes out the model in terms of a ***baseline group*** and deviations from that baseline or reference level. The reference-coded model for the $i^{th}$ subject in the $j^{th}$ group is $y_{ij} = {\color{purple}{\boldsymbol{\alpha + \tau_j}}}+\varepsilon_{ij}$ where $\color{purple}{\boldsymbol{\alpha}}$ ("alpha") is the true mean for the baseline group (usually first alphabetically) and the $\color{purple}{\boldsymbol{\tau_j}}$ (tau $j$) are the deviations from the baseline group for group $j$. The deviation for the baseline group, $\color{purple}{\boldsymbol{\tau_1}}$, is always set to 0 so there are really just deviations for groups 2 through $J$. The equivalence between the reference-coded and cell means models can be seen by considering the mean for the first, second, and $J^{th}$ groups in both models:

$$\begin{array}{lccc}
& \textbf{Cell means:} && \textbf{Reference-coded:}\\
\textbf{Group } 1: & \color{red}{\mu_1} && \color{purple}{\boldsymbol{\alpha}} \\
\textbf{Group } 2: & \color{red}{\mu_2} && \color{purple}{\boldsymbol{\alpha + \tau_2}} \\
\ldots & \ldots && \ldots \\
\textbf{Group } J: & \color{red}{\mu_J} && \color{purple}{\boldsymbol{\alpha +\tau_J}}
\end{array}$$

> -- <cite>https://greenwood-stat.github.io/GreenwoodBookHTML/chapter3.html#section3-2</cite>

Further reading:

[A guide to creating design matrices for gene expression experiments](https://pmc.ncbi.nlm.nih.gov/articles/PMC7873980/)

[Linear models from bioinformatics-core-shared-training.github.io](https://bioinformatics-core-shared-training.github.io/Bulk_RNAseq_Course_June24/Bulk_RNAseq_Course_Base/Markdowns/07_Linear_Models.html#Common_Designs)

[STAT 502 Analysis of Variance and Design of Experiments from Penn State](https://online.stat.psu.edu/stat502/lesson/welcome-stat-502)

Here we use the model formula as `~ kit + ilb_trt`.

```{r, Fit model}
dds <- DESeqDataSetFromMatrix(countData = count_df,
                              colData = meta,
                              design = ~ + ilb_trt + kit)

## filter gene with less than 10 reads
dds <- dds[ rowSums(counts(dds)) > 10, ]

dds <- DESeq(dds)
res <- results(dds)

print("model matix")
model.matrix(design(dds), colData(dds))

print("ExploreModelMatrix")
vd <- VisualizeDesign(sampleData = meta,
                      designFormula = ~ ilb_trt + kit,
                      textSizeFitted = 4)

cowplot::plot_grid(plotlist = vd$plotlist)
cowplot::plot_grid(plotlist = vd$cooccurrenceplots)

# coefficients
print("model coefficients")
resultsNames(dds)

# fdr filtering
# res <- results(dds, contrast=c("ilb_trt","treated","untreated"), alpha=0.1)

## shrink logFC
res <- lfcShrink(dds, contrast=c("ilb_trt","treated","untreated"), type="ashr", lfcThreshold=1)
res <- res[order(res$pvalue), ]
## MA plot
# plotMA(res)

out_df <- as.data.frame(res) |>
  tibble::rownames_to_column(var = 'id') |>
  left_join(annot_df, by = 'id') |>
  dplyr::rename(gene_symbol = geneSymbol,
                gene_id = id) |>
  mutate(uniq_symbol = scater::uniquifyFeatureNames(ID = gene_id, names = gene_symbol)) |>
  relocate(gene_symbol, .after = gene_id) |>
  relocate(uniq_symbol, .after = gene_symbol) |>
  as_tibble()

## create a res output file
# write_tsv(out_df, "somepath")
```

## Top 20 genes heatmap using rlog

```{r, Top 20 genes heatmap, fig.width=6, fig.height=6}
top_genes <- out_df |>
  arrange(padj) |>
  # dplyr::arrange(dplyr::desc(abs(log2FoldChange))) |>
  head(20) |>
  dplyr::select(gene_id, uniq_symbol) |>
  as.data.frame() |>
  as_tibble()
  # tibble::column_to_rownames(var = "uniq_symbol")

mat <- assay(rld_filt) |>
  as.data.frame() |>
  tibble::rownames_to_column(var = "gene_id") |>
  as_tibble()

mat <- left_join(top_genes, mat, by = "gene_id") |>
  as.data.frame() |>
  dplyr::select(-gene_id) |>
  tibble::column_to_rownames(var = 'uniq_symbol')

annotation_col = data.frame(
    ilb_trt = meta$ilb_trt
    # kit = meta$kit
)

set.seed(0)
hmap <- pheatmap(mat,
                # clustering_distance_columns = sampleDists,
                # clustering_distance_rows = sampleDists,
                # rect_gp = gpar(col = "white", lwd = 2),
                # # row_names_gp = gpar(fontsize = 5),
                # # row_dend_side = 'left',
                # column_title = "Heatmap of the sample-to-sample based on Euclidean distances",
                heatmap_legend_param = list(
                                title = "centered rlog"),
                annotation_col = annotation_col
                )

print(hmap)
```

## Volcano plot

```{r, Volcano plot, fig.width=4, fig.height=4}}
make_volcano <- function(df, pval_nm, pval_cutoff=0.1){
  # remove genes with NA for pvalue

  df <- df |> tidyr::drop_na(pval_nm)

  top_genes <- df |>
    dplyr::arrange(desc(abs(df$log2FoldChange))) |>
    dplyr::filter(row_number() <= 10) |>
    dplyr::select(uniq_symbol) |>
    as.vector()

  df$Sig <- ifelse(df$padj <= pval_cutoff, "Sig", "NS")

  df[[pval_nm]] <- -log10(df[[pval_nm]])

  ggplot(df, aes(x=log2FoldChange, y=.data[[pval_nm]])) +
    geom_point(aes(color=Sig), size=0.6) +
    scale_color_manual(values=c("black", "salmon")) +
    theme_bw() + ylab(stringr::str_glue("-log10(", pval_nm,")")) +
    ggrepel::geom_text_repel(data=df |> dplyr::filter(uniq_symbol %in% top_genes$uniq_symbol),
                    aes(label=uniq_symbol), max.overlaps=Inf, min.segment.length = 0) +
    ggprism::scale_color_prism() +
    ggprism::scale_fill_prism()
}

make_volcano(out_df, pval_nm="padj", pval_cutoff=0.05)
```

## Play around with contrast vectors

```{r, Contrast Vector}
print("model coefficients")
resultsNames(dds)

print("model matrix - truseq and untreated are the refence levels")
model.matrix(design(dds), colData(dds))

untrt_truseq <- c(1,0,0,0)
trt_truseq <- c(1,1,0,0)
untrt_pico <- c(1,0,1,0)
untrt_v4 <- c(1,0,0,1)

results(dds, contrast = trt_truseq - untrt_truseq)
results(dds, contrast = list('ilb_trt_treated_vs_untreated'))

results(dds, contrast = list('kit_v4_vs_truseq'))
results(dds, contrast = untrt_v4 - untrt_truseq)
```
